Pattern matching for\\ $(213,231)$-avoiding permutations

### 2015-06-18 article.tex
LNCS 12 pages + appendix

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	(The algorithm also improves on the previous
	$O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
	(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for LCS-like issues
	of separable permutations).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers
	(sequence A006318 in OEIS).
	To the best of our knowledge,
	separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

@@

	We focus in this paper on pattern matching issues for
	$(213,231)$-avoiding permutations
    (\emph{i.e.}, those permutations that avoid both $213$ and $231$).
	The number of $n$-permutations that avoid both
	$213$ and $231$ is
	$t_0 = 1$ for $n = 0$ and
	$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
	On an individual basis,
	the permutations that do not contain the permutation pattern $231$
	are exactly the \emph{stack-sortable permutations} and they are counted by
	the Catalan numbers \cite{Knuth:1997:ACP:260999}.
	A stack-sortable permutation is a permutation whose elements may be sorted by
	an algorithm whose internal storage is limited to a single stack data structure.
	As for $213$, it is well-known that
 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
 	the reverse of $\pi'$ avoids $213$.
 	From a combinatorial point of view,
 	B\'ona \cite{Bona:ElJC:2012}
 	showed the rather surprising fact that the cumulative number of
 	occurrences of the classical patterns $231$ and $213$ are the same on the
 	set of permutations avoiding $132$,
 	beside the pattern based statistics $231$ and $213$
 	do not have the same distribution on this set.
 	\emph{Almost avoidance} has also been considered;
 	a permutation $\pi$ almost avoids a set $X$ of permutations
 	if there is a way to remove a single element of $\pi$ to get a permutation
 	that avoids all elements in $P$
 	and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
 	$P$.
 	It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
 	$L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.

	This paper is organized as follows.
	In Section~\ref{section:Definitions} the needed definition are presented.
	Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
	an online linear-time algorithm in case both the pattern and the target
	permutation are $(213,231)$-avoiding,
	whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
	where the pattern only is $(213,231)$-avoiding.
	In Section~\ref{section:bivincular} we give a polynomial-time algorithm
	for $(213,231)$-avoiding bivincular patterns.
	In Section~\ref{section:LCS} we consider the problem of finding the longest
	$(213,231)$-avoiding pattern in a permutation.
	In Section~\ref{section:LCS common 2 permutations} we give a
	polynomial-time algorithm for computing the longest
	$(213,231)$-avoiding pattern common to two permutations.

@@

A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$
if there exists a subsequence of (not necessarily consecutive)
entries of $\pi$ that has the same relative order as $\sigma$,
and in this case $\sigma$ is said to be a \emph{pattern} of
$\pi$, written $\sigma \preceq \pi$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example, the permutation $\pi = 391867452$
contains the pattern $\sigma = 51342$,
as can be seen in the highlighted subsequence of
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}\mathbf{4}52$
(or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}4\mathbf{5}2$
or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}45\mathbf{2}$
).
Each subsequence $91674$,
$91675$,
$91672$ in $\pi$ is called a
\emph{copy},
(or \emph{instance} or \emph{occurrence}) of $\sigma$.
Since the permutation $\pi = 391867452$  contains no increasing subsequence of
length four, $\pi$ avoids $1234$.

@@

\begin{proof}
Any other initial element would serve as a `$2$' in either a
$231$ or $213$ with $1$ and $n$ as the `$1$' and `$3$' respectively.
\qed
\end{proof}

@@

\begin{corollary}
\label{corollary:minmaxelement}
$\pi \in \AV_n(213,231)$ if and only if for $1 \leq i \leq n$,
$\pi[i]$ is the minimal or the maximal element of $\pi[i:]$.
\end{corollary}

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi \in \AV_n(213,231)$ and $1 \leq i < n$. Then,
(1) 
if $\pi[i]$ is an ascent element then $\pi[i]$ is the minimal element 
of $\pi[i:]$
and 
(2)
if if $\pi[i]$ is a descent element then $\pi[i]$ is the maximal element 
of $\pi[i:]$
\end{corollary}

@@

\begin{proof}
  The forward direction is obvious.
  We prove the backward direction by induction on the size of the motif
  $\sigma$.
  The base case is a motif of size $2$.
  Suppose first that $\sigma = 12$ and hence $s_\sigma = a$.
  Let $t = \pi_{i_1}\pi_{i_2}$, $i_1 < i_2$, be a subsequence of $\pi$
  such that $s_t$ occurs as a subsequence in $s_\pi$.
  Since $s_\sigma = s_t = a$, this reduces to saying that
  $\pi_{i_1} < \pi_{i_2}$, and hence $t$ is an occurrence of $\sigma = 12$ in $\pi$.
  A similar argument shows that the lemma holds true for $\sigma = 21$.
  Now, assume that the lemma is true for all motifs up to size $k \geq 2$.
  Let $\sigma \in \Av{k+1}{231,213}$ and
  let $t = \pi_{i_1}\pi_{i_2}\,\ldots\,\pi_{i_{k+1}}$,
  $i_1 < i_2 < \ldots < i_{k+1}$,
  be a subsequence of $\pi$ of length $k+1$ such that
  $s_t = s_\sigma$.
  Suppose that $s_t$ occurs in $s_\pi$ as a subsequence.
  Then, $s_t[2:] = s_\sigma[2:]$ occurs as a subsequence in $s_\pi[2:]$.
  But $\sigma[2:] \in \Av{k}{231,213}$ and hence,
  by the inductive hypothesis, it follows that
  $t[2:]$ is an occurrence of $\sigma[2:]$ in $\pi[2:]$.
  Moreover, if $s_\sigma[1] = a$ (resp. $s_\sigma[1] = d$)
  then $\sigma[1]$ is the minimal (resp. maximal) element of $\sigma$
  (Lemma~\ref{lemma:first element is 1 or n}),
  and hence, since $s_t = s_\sigma$ occurs as a subsequence in $s_\pi$,
  $t[1]$ is the minimal (resp. maximal) element of $t$.
  Therefore, $t$ is an occurrence of $\sigma$ in $\pi$.
  \qed
\end{proof}

@@

\begin{proof}
According to Lemma~\ref{lemma:MatchStripeToPermutation} the problem reduces
to deciding whether $s_\sigma$ occurs as a subsequence in $s_\pi$.
A natural greedy approach solves this issue in linear-time.
\qed
% To solve the permutation pattern for a permutation in $\AV_n(213,231)$,
% we find a find a match of the stripe of the pattern in the stripe of the text.
% This can be done linearly by a greedy algorithm,
% by matching a step whenever it is possible.
% And thank to the definition of the definition of a stripe in
% Corollary~\ref{corollary:minmaxelement},
% we do not have to compute the stripe in advance.
% Therefore we have an linear (online) algorithm.
\end{proof}

Observe that, according to Corollary~\ref{corollary:minmaxelement},
the effective construction of the two stripes is not required in stretch 
of algorithm given for the proof in the proposition \ref{Proposition:both permutations are avoiding},
and hence we have an online algorithm.

@@

\begin{proof}
Consider the rightmost element of $\factor{\sigma}{i}$ and assume this element is an ascent.
Then, it follows that it is smaller than the following element which is $\first{\sigma}{i-1}$.
Now, since $\factor{\sigma}{i}$ is an increasing subsequence, every element of
$\factor{\sigma}{i}$ is smaller than $\first{\sigma}{i-1}$.
According to Lemma\ref{lemma:first element is 1 or n}, $\first{\sigma}{i-1}$ is the maximal (minimal) element of $\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$, and hence
$\first{\sigma}{i-1}$ is the maximal  element of
$\factor{\sigma}{i}\factor{\sigma}{i-1}\ldots\factor{\sigma}{1}$.

we now define the set
$\SET{\pmotif}{\ptext}{i}{j}$ as follows:
given a subsequence $s$ of $\ptext[j:]$,
$s \in \SET{\pmotif}{\ptext}{i}{j}$ if and only if
$s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$
and the element $\first{\pmotif}{i}$ occurs in $\ptext[j]$.

\begin{lemma}
Let $\pmotif$ be a permutation , 
$\factor{\pmotif}{i}$ be an ascent (resp. descent) factor,
$s$ be a subsequence such that $s \in \SET{\pmotif}{\ptext}{i}{j}$ and $s[\first{\pmotif}{i-1}]$ is minimal (resp. maximal). 
For all subsequences 
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and for all subsequences $t$ of $\ptext$, such that $t=t's'$, if $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ then the subsequence $t's$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ in $\ptext[j:]$ such that $\first{\pmotif}{i}$ occurs in $\ptext[j]$.
\end{lemma}

\begin{proof}
By definition $s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$. To prove that $ts$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ we prove that every element of t is larger than every element of $s$. If $ts'$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ so every element of $t$ is larger than every element of $s'$. Moreover the maximal element of $s$ is smaller than  the maximal element of $s'$ so every element of $s$ is smaller than every element of $s'$ thus every element of $s$ is smaller than every element of $t$. We use a similar argument if $\factor{\pmotif}{i}$ a descent factor.
\qed
\end{proof}

\begin{corollary}
Let $\pmotif$ be a permutation , 
$\factor{\pmotif}{i}$ be an ascent (resp. descent) factor
and 
$s$ be a subsequence such that $s \in \SET{\pmotif}{\ptext}{i}{j}$ and $s[\first{\pmotif}{i-1}]$ is minimal (resp. maximal). 
These following statements are equivalent :
\begin{itemize}
	\item There exists an 
	occurrence of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ occurs in $\ptext[j]$.
	\item There exists a subsequence $t$ of $\ptext[1:j-1]$  such that $ts$ is an occurrence of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ occurs in $\ptext[j]$. 
\end{itemize}
\end{corollary}

@@

% For the proof and the algorithm we introduce a decomposition : we decompose the stripe of the motif into factor of same letter, and label those factor from right to left. We also need to add the last index into the first factor. For each factor, we denote the left most element by $\firsta$. For example, given the motif $981237654$, $\factor{1} =7654$, $\first{1}=7$, $\factor{2}=123$, $\first{2}=1$, $\factor{3}=98$ and $\first{9}$.

\begin{proof}%[of Proposition~\ref{Proposition:sigma avoids 213 and 231}]
%Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, the safest match is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. Indeed when we will match $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$, every element of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$, the optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element at index $j'$ of $\ptext[j:]$ such that : (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1).
%
%\begin{remark}
%Given an ascent (descent) factor $\factor{i}$, the maximal (minimal) element of $\factor{i}\factor{i-1}\cdots\factor{1}$ is $\first{i-1}$.
%\end{remark}
%
%\begin{remark}
%An ascent (descent) factor 	$\factor{i}$ is a match in $\ptext[j:j']$  if and only if there exists an increasing (decreasing) subsequence in $\ptext[j:j']$ of size superior or equal to the size of $\factor{i}$.
%\end{remark}

Consider the following problem :
$$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}

@@

\noindent\textbf{BASE :}
\begin{align*}
\LM{\pmotif}{\ptext}{1}{j} 
&=
\begin{cases}
		\min_{j<j'}\{\infty\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LIS(j,j',\ptext[j'])  $}\}&\text{ an ascent element } \\
		&\\
		\max_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LDS(j,j',\ptext[j']) $}\}&\text{ a descent element }\\
\end{cases}
\\
\intertext{
In the base case,
one is looking for an occurrence of the first factor.
If the factor has only ascent (resp. descent) elements, then
one has to find an increasing (resp. decreasing) subsequence
in the text of the same size or that is longer that
the size of the first factor. To find the occurrence that is
the less restrictive (for the future matching)
one must find the occurrence with the right most element
of the sequence minimal (resp. maximal).}
\\
\intertext{\textbf{STEP :}}
\LM{\pmotif}{\ptext}{i}{j} 
&=
\begin{cases}
	\min \{\infty\} \cup  \AF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is an ascent element}\\
	\max \{0\} \cup  \DF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is  a descent element}\\
\end{cases}
\end{align*}
where $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ are the sets 
of elements such that if  
$\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$ or if $\ptext[k] \in \DF{\pmotif}{\ptext}{i}{j}$ 
then there exists an occurrence $t$ of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element 
$\first{\pmotif}{i-1}$ occurs in element $\ptext[k]$. 
$t$ exists if and only if $t=t's'$ with $t'$ is an occurrence of  $\factor{\pmotif}{i}$ in $\ptext[j:k-1]$ with element
$\first{\pmotif}{i-1}$ occurs in element $\ptext[k]$
and $s \in \SET{\pmotif}{\ptext}{i-1}{k}$.
If $\factor{\pmotif}{i}$ is an increasing (resp. decreasing)
sequence $t'$ must be an increasing (resp. decreasing) 
subsequence
in $\ptext[j:k-1]$ of the same size or that is longer that
the size of $\factor{\pmotif}{i}$ (denoted $|\factor{\pmotif}{i}|$).
Moreover one must assure that the elements of subsequence 
$t'$ are smaller (resp. bigger) than the element of $s'$ 
in particular the minimal (resp. maximal) element.
Thus we can compute $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ as follows:
\begin{align*}
\AF{\pmotif}{\ptext}{i}{j}
&=
\text{$\{\ptext[j'] \;|\; j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq 0$ and} 
\\
&\qquad 
\text{$|\factor{\pmotif}{i}| \leq LIS_{\ptext}(j,j'-1,\LM{\pmotif}{\ptext}{i-1}{j'})\}$} 
\\
\DF{\pmotif}{\ptext}{i}{j}
&=
\text{$\{\ptext[j'] \;|\; j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq \infty$ and}
\\
&\qquad
\text{$|\factor{\pmotif}{i}| \leq LDS_{\ptext}(j,j'-1,\LM{\pmotif}{\ptext}{i-1}{j'})\}$}
\end{align*}
where $LIS_{\ptext}(i,j,k)$ (resp. $LDS_{\ptext}(i,j,k)$) is the longest increasing 
(resp. decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
smaller (resp. bigger) than $k$.
$LIS_{\ptext}$ and $LDS_{\ptext}$ can be computed in 
$O(|\ptext|^2\log(\log(|\ptext|)))$ time (see \cite{Bespamyatnikh00enumeratinglongest}).

@@

\begin{proof}
Suppose that there exists $k$, $i<k<j$,
such that $\pmotif[k]$ is ascent. Ascent elements are increasing  so $\sigma[i]<\sigma[k]<\sigma[j]$ which is in contradiction with $\sigma[j]=\sigma[i]+1$.
We use a similar argument if $\sigma[i]$ is a descent element
\qed
\end{proof}

@@

\noindent\textbf{STEP :}

>>>>

	A few particular cases of the pattern containment problem have been attacked successfully.
	The case of increasing patterns is solvable in
	$O(|\pi| \log \log |\sigma|)$ time in the RAM model \cite{Crochemore:Porat:2010},
	improving the previous 30-year bound of $O(|\pi| \log |\sigma|)$.
	% (The algorithm also improves on the previous
	% $O(|\pi| \log \log |\pi|)$ bound.)
	Furthermore, the patterns $132$, $213$, $231$, $312$ can all be handled in linear-time
	by stack sorting algorithms.
	Any pattern of length $4$ can be detected in $O(|\pi| \log |\pi|)$ time
	\cite{Albert:Aldred:Atkinson:Holton:ISAAC:2001}.
	Algorithmic issues for $321$-avoiding patterns containment has been investigated in
	\cite{Guillemot:Vialette:ISAAC:2009}.
	The pattern containment problem is also solvable in
	polynomial-time for separable patterns \cite{Ibarra:1997,Bose:Buss:Lubiw:1998}
	(see also \cite{Bouvel:Rossin:Vialette:CPM:2007} for LCS-like issues
	of separable permutations).
	Separable permutations are those permutations that contain neither
	$2413$ nor $3142$, and they are enumerated by the Schr{\"o}der numbers
	(sequence A006318 in OEIS).
	% To the best of our knowledge,
	Separable permutations first arose in the work of
	Avis and Newborn~\cite{Avis:Newborn:1981},
	who showed that they are precisely the permutations which can be sorted by an
	arbitrary number of pop-stacks in series,
	where a pop-stack is a restricted form of stack in which any pop operation
	pops all items at once.
	(Notice that the separable permutations include as a special case the
	stack-sortable permutations, which avoid the pattern $231$.)

@@

	We focus in this paper on pattern matching issues for
	$(213,231)$-avoiding permutations
    (\emph{i.e.}, those permutations that avoid both $213$ and $231$).
	The number of $n$-permutations that avoid both
	$213$ and $231$ is
	$t_0 = 1$ for $n = 0$ and
	$t_n =2^{n-1}$ for $n\geq 1$ \cite{Simion:Schmidt:EJC:1985}.
	On an individual basis,
	the permutations that do not contain the permutation pattern $231$
	are exactly the \emph{stack-sortable permutations} and they are counted by
	the Catalan numbers \cite{Knuth:1997:ACP:260999}.
	A stack-sortable permutation is a permutation whose elements may be sorted by
	an algorithm whose internal storage is limited to a single stack data structure.
	As for $213$, it is well-known that
 	if $\pi = \pi_1\pi\,\ldots\,\pi_n$ avoids $132$, then its complement
 	$\pi' = (n+1-\pi_1)(n+1-\pi_2)\,\ldots\,(n+1-\pi_n)$ avoids $312$, and
 	the reverse of $\pi'$ avoids $213$.
 	% From a combinatorial point of view,
 	% B\'ona \cite{Bona:ElJC:2012}
 	% showed the rather surprising fact that the cumulative number of
 	% occurrences of the classical patterns $231$ and $213$ are the same on the
 	% set of permutations avoiding $132$,
 	% beside the pattern based statistics $231$ and $213$
 	% do not have the same distribution on this set.
 	% \emph{Almost avoidance} has also been considered;
 	% a permutation $\pi$ almost avoids a set $X$ of permutations
 	% if there is a way to remove a single element of $\pi$ to get a permutation
 	% that avoids all elements in $P$
 	% and $L_n(P)$ denotes the set of permutations of length $n$ that almost avoid
 	% $P$.
 	% It is shown in \cite{Griffiths:Smith:Warren:PMA:2011} that
 	% $L_n(213, 231) = L_n(132, 231) = L_n(132, 312) = L_n(213, 312)$.

	This paper is organized as follows.
	In Section~\ref{section:Definitions} the needed definition are presented.
	Section~\ref{section:both are (213,231)-avoiding} is devoted to presenting
	an online linear-time algorithm in case both the pattern and the target
	permutation are $(213,231)$-avoiding,
	whereas Section~\ref{section:sigma only avoid 231 and 213} focuses on the case
	where the pattern only is $(213,231)$-avoiding.
	In Section~\ref{section:bivincular} we give a polynomial-time algorithm
	for $(213,231)$-avoiding bivincular patterns.
	In Section~\ref{section:LCS} we consider the problem of finding the longest
	$(213,231)$-avoiding pattern that occurs in one or two permutations.

@@

A permutation $\pi$ is said to \emph{contain} the permutation $\sigma$
if there exists a subsequence of (not necessarily consecutive)
entries of $\pi$ that has the same relative order as $\sigma$,
and in this case $\sigma$ is said to be a \emph{pattern} of
$\pi$, written $\sigma \preceq \pi$.
Otherwise, $\pi$ is said to \emph{avoid} the permutation $\sigma$.
For example, the permutation $\pi = 391867452$
contains the pattern $\sigma = 51342$,
as can be seen in the highlighted subsequence of
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}\mathbf{4}52$
(or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}4\mathbf{5}2$
or
$\pi = 3\mathbf{9}\mathbf{1}8\mathbf{6}\mathbf{7}45\mathbf{2}$
).
Each subsequence $91674$,
$91675$,
$91672$ in $\pi$ is called a
\emph{copy},
(or \emph{instance} or \emph{occurrence}) of $\sigma$.
% Since the permutation $\pi = 391867452$  contains no increasing subsequence of
% length four, $\pi$ avoids $1234$.

@@

@@

\begin{corollary}
\label{corollary:minmaxelement}
Let $\pi \in S_n$. 
Then,
$\pi \in \AV_n(213,231)$ if and only if for $1 \leq i \leq n$
$\pi[i]$ is the minimal or the maximal element of $\pi[i:]$.
\end{corollary}

\begin{corollary}
\label{corollary:max is ascent}
Let $\pi \in \AV_n(213,231)$ and $1 \leq i < n$. Then,
if $\pi[i]$ is an ascent (resp. descent) element 
then $\pi[i]$ is the minimal (resp. maximal) element 
of $\pi[i:]$.
\end{corollary}

@@

@@

Observe that, according to Corollary~\ref{corollary:minmaxelement},
the effective construction of the two stripes is not required in stretch 
of algorithm given for the proof in 
Proposition~
\ref{Proposition:both permutations are avoiding},
and hence we have an online algorithm.

@@

We now define the set
$\SET{\pmotif}{\ptext}{i}{j}$ as follows:
given a subsequence $s$ of $\ptext[j:]$,
$s \in \SET{\pmotif}{\ptext}{i}{j}$ if and only if
$s$ is an occurrence of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$
and the element $\first{\pmotif}{i}$ occurs in $\ptext[j]$.

\begin{lemma}
\label{lemma:ts}
Let $\pmotif$ be a permutation , 
$\factor{\pmotif}{i}$ be an ascent (resp. descent) factor,
$s$ be a subsequence such that $s \in \SET{\pmotif}{\ptext}{i}{j}$ and $s[\first{\pmotif}{i-1}]$ is minimal (resp. maximal). 
For all subsequences 
$s' \in \SET{\pmotif}{\ptext}{i}{j}$ and for all subsequences $t$ of $\ptext$, such that $t=t's'$, if $t$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ then the subsequence $t's$ is an occurrence of $\pmotif[\firsti{\pmotif}{i+1}:]$ in $\ptext[j:]$ such that $\first{\pmotif}{i}$ occurs in $\ptext[j]$.
\end{lemma}

\begin{corollary}
Let $\pmotif$ be a permutation , 
$\factor{\pmotif}{i}$ be an ascent (resp. descent) factor
and 
$s$ be a subsequence such that $s \in \SET{\pmotif}{\ptext}{i}{j}$ and $s[\first{\pmotif}{i-1}]$ is minimal (resp. maximal). 
These following statements are equivalent:
(1)
There exists an 
occurrence of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ occurs in $\ptext[j]$,
and (2)
There exists a subsequence $t$ of $\ptext[1:j-1]$  such that $ts$ is 
an occurrence of $\pmotif$ in $\ptext$ with element $\first{\pmotif}{i}$ occurs 
in $\ptext[j]$. 
\end{corollary}

@@

% For the proof and the algorithm we introduce a decomposition: we decompose the stripe of the motif into factor of same letter, and label those factor from right to left. We also need to add the last index into the first factor. For each factor, we denote the leftmost element by $\firsta$. For example, given the motif $981237654$, $\factor{1} =7654$, $\first{1}=7$, $\factor{2}=123$, $\first{2}=1$, $\factor{3}=98$ and $\first{9}$.

\begin{proof}%[of Proposition~\ref{Proposition:sigma avoids 213 and 231}]
%Given an ascent (descent) factor $\factor{i}$ of a motif $\pmotif$, if we want to match the suffix of $\pmotif$ starting by $\factor{i}$, $\factor{i}\factor{i-1}\cdots\factor{1}$ and given that we know every match of $\factor{i-1}\cdots\factor{1}$, the safest match is to choose the match of $\factor{i-1}\cdots\factor{1}$ that minimize (maximize) the maximal (minimal) element of the match of $\factor{i-1}\cdots\factor{1}$ and that allow a match of factor $\factor{i}$. Indeed when we will match $\factor{i+1}\factor{i}\factor{i-1}\cdots\factor{1}$, every element of $\factor{i+1}$ must be matched to element superior (inferior) to every element matched element of $\factor{i}\factor{i-1}\cdots\factor{1}$, particularly its must be superior (inferior) to the maximal (minimal) element of the match of $\factor{i}\factor{i-1}\cdots\factor{1}$. In other word if we want to match $\pmotif[\first{i}:]$ with $\ptext[j:]$ such that $\first{i}$ is matched with element at index $j$ in $\ptext$, the optimal way of matching $\pmotif[\first{i-1}:]$ is to choose the minimal (maximal) element at index $j'$ of $\ptext[j:]$ such that: (1) $\pmotif[\first{i-1}:]$ is matched with $\ptext[j':]$ and element $\first{i-1}$ is matched with element at index $j'$, (2) the factor $\factor{i}$ is matched with $\ptext[j,j'-1]$ and element $\first{i}$ is matched with element at index $j$, (3) Every elements of the match of (2) if inferior (superior) to the minimal (maximal) element of the match of (1).
%
%\begin{remark}
%Given an ascent (descent) factor $\factor{i}$, the maximal (minimal) element of $\factor{i}\factor{i-1}\cdots\factor{1}$ is $\first{i-1}$.
%\end{remark}
%
%\begin{remark}
%An ascent (descent) factor 	$\factor{i}$ is a match in $\ptext[j:j']$  if and only if there exists an increasing (decreasing) subsequence in $\ptext[j:j']$ of size superior or equal to the size of $\factor{i}$.
%\end{remark}

Consider the following problem:
$$
\LM{\pmotif}{\ptext}{i}{j} =
\begin{cases}

@@

\noindent\textbf{BASE:}
\begin{align*}
\LM{\pmotif}{\ptext}{1}{j} 
&=
\begin{cases}
		\min_{j<j'}\{\infty\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LIS(j,j',\ptext[j'])  $}\}&\text{ an ascent element } \\
		&\\
		\max_{j<j'}\{0\} \cup \{\ptext[j'] |\text{ $j'$ such that \indent } 
			& \text{If $\first{\sigma}{i}$ is}  \\
		\text{ $|\factor{\pmotif}{1}| \leq LDS(j,j',\ptext[j']) $}\}&\text{ a descent element }\\
\end{cases}
\\
\intertext{
In the base case,
one is looking for an occurrence of the first factor.
If the factor has only ascent (resp. descent) elements, then
one has to find an increasing (resp. decreasing) subsequence
in the text of the same size or that is longer that
the size of the first factor. To find the occurrence that is
the less restrictive (for the future matching)
one must find the occurrence with the rightmost element
of the sequence minimal (resp. maximal).}
\\
\intertext{\textbf{STEP:}}
\LM{\pmotif}{\ptext}{i}{j} 
&=
\begin{cases}
	\min \{\infty\} \cup  \AF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is an ascent element}\\
	\max \{0\} \cup  \DF{\pmotif}{\ptext}{i}{j} &
	\text{If $\first{\sigma}{i}$ is  a descent element}\\
\end{cases}
\end{align*}
where $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ are the sets 
of elements such that if  
$\ptext[k] \in \AF{\pmotif}{\ptext}{i}{j}$ or if $\ptext[k] \in \DF{\pmotif}{\ptext}{i}{j}$ 
then there exists an occurrence $t$ of $\pmotif[\firsti{\pmotif}{i}:]$ in $\ptext[j:]$ and the element 
$\first{\pmotif}{i-1}$ occurs in element $\ptext[k]$. 
$t$ exists if and only if $t=t's'$ with $t'$ is an occurrence of  $\factor{\pmotif}{i}$ in $\ptext[j:k-1]$ with element
$\first{\pmotif}{i-1}$ occurs in element $\ptext[k]$
and $s \in \SET{\pmotif}{\ptext}{i-1}{k}$.
If $\factor{\pmotif}{i}$ is an increasing (resp. decreasing)
sequence $t'$ must be an increasing (resp. decreasing) 
subsequence
in $\ptext[j:k-1]$ of the same size or that is longer that
the size of $\factor{\pmotif}{i}$ (denoted $|\factor{\pmotif}{i}|$).
Moreover one must assure that the elements of subsequence 
$t'$ are smaller (resp. bigger) than the element of $s'$ 
in particular the minimal (resp. maximal) element.
Thus we can compute $\AF{\pmotif}{\ptext}{i}{j}$ and $\DF{\pmotif}{\ptext}{i}{j}$ as follows:
\begin{align*}
\AF{\pmotif}{\ptext}{i}{j}
&=
\text{$\{\ptext[j'] \;|\; j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq 0$ and} 
\\
&\qquad 
\text{$|\factor{\pmotif}{i}| \leq LIS_{\ptext}(j,j'-1,\LM{\pmotif}{\ptext}{i-1}{j'})\}$} 
\\
\DF{\pmotif}{\ptext}{i}{j}
&=
\text{$\{\ptext[j'] \;|\; j<j'$ and $\LM{\pmotif}{\ptext}{i-1}{j'} \neq \infty$ and}
\\
&\qquad
\text{$|\factor{\pmotif}{i}| \leq LDS_{\ptext}(j,j'-1,\LM{\pmotif}{\ptext}{i-1}{j'})\}$}
\end{align*}
where $LIS_{\ptext}(i,j,k)$ (resp. $LDS_{\ptext}(i,j,k)$) is the longest increasing 
(resp. decreasing) sequence in $\ptext$ starting at $i$ and ending at $j$,
with every element of this sequence
smaller (resp. bigger) than $k$.
$LIS_{\ptext}$ and $LDS_{\ptext}$ can be computed in 
$O(|\ptext|^2\log(\log(|\ptext|)))$ time (see \cite{Bespamyatnikh00enumeratinglongest}).

@@

@@

\noindent\textbf{STEP:}
